Parameter Inference\\with Bifurcation Diagrams

### 2021-05-26 docs/article.tex

\subsection{Genetic Toggle Switch}
In this section we optimise a model where the states share a Hill function relationship with cooperatively $n=2$; these models often emerge from mass action kinetics with quasi-steady state approximations and are used to model species concentrations. After re-scaling the equations governing the dynamics of concentrations, the simplified equations for state $u_1$ and $u_2$ become 
\begin{equation}
    \partial_t u_1 = \frac{ a_1 + (p u_2)^2}{ 1 + (p u_2)^2 } - \mu_1 u_1 \quad
    \partial_t u_2 = \frac{ a_2 + (k u_1)^2}{ 1 + (k u_1)^2 } - \mu_2 u_2
    \label{eq:two-state}
\end{equation}
where $a_k$ is the baseline production rate for species $k$ in the absence of the other species. Each species has a finite degradation rate $\mu_k$. Finally we have two half-saturation constants $p$ and $k$, one of which is chosen as our control condition. A baseline production rate $a_k>1$ recovers an inhibitor type hill function for species $k$ and is an activator otherwise. The half-saturation are proportional to the slope of the hill productions. Solving for the steady states,  substituting the equation for $u_1$ into $u_2$ and rearranging gives rise to the relationship
\begin{equation}
    \dfrac{k}{\mu_1} = \dfrac{(1 + (p u_2)^2) \sqrt{a_2 - \mu_2 u_2} }{ (a_1 + (p u_2)^2) \sqrt{ \mu_2 u_2 - 1 } }
    \label{eq:steady-states}
\end{equation}
which reveals that only the ratio between half-saturation and degradation $\frac{k}{\mu_1}$ affects the solutions to this equation, and hence the locations of the bifurcations. Figure \ref{fig:two-state-optima} reveals the results for 800 optimisation runs with 200 ADAM iterations. 98\% of runs converged onto one of the two clusters: mutual activation and mutual inhibition regimes, examples of which are shown as insets.
\begin{figure}
\centering
%\includegraphics[width=11cm]{two-state-optima}
\includegraphics[width=11cm]{two-state-optima.png}
\caption{Optimal parameter estimates $\theta^*$ for the two state model \eqref{eq:two-state} for the targets $\targets=\{4,5\}$ reveal two clusters of qualitatively different regimes: mutual activation and mutual inhibition. }
\label{fig:two-state-optima}
\end{figure}

>>>>

\subsection{Genetic Toggle Switch}
In this section we optimise a model where the states share a Hill function relationship with cooperatively $n=2$; these models often emerge from mass action kinetics with quasi-steady state approximations and are used to model species concentrations. After re-scaling the equations governing the dynamics of concentrations, the simplified equations for state $u_1$ and $u_2$ become 
\begin{equation}
    \partial_t u_1 = \frac{ a_1 + (p u_2)^2}{ 1 + (p u_2)^2 } - \mu_1 u_1 \quad
    \partial_t u_2 = \frac{ a_2 + (k u_1)^2}{ 1 + (k u_1)^2 } - \mu_2 u_2
    \label{eq:two-state}
\end{equation}
where $a_k$ is the baseline production rate for species $k$ in the absence of the other species. Each species has a finite degradation rate $\mu_k$. Finally we have two half-saturation constants $p$ and $k$, one of which is chosen as our control condition. A baseline production rate $a_k>1$ recovers an inhibitor type hill function for species $k$ and is an activator otherwise. The half-saturation are proportional to the slope of the hill productions. Solving for the steady states,  substituting the equation for $u_1$ into $u_2$ and rearranging gives rise to the relationship
\begin{equation}
    \dfrac{k}{\mu_1} = \dfrac{(1 + (\frac{p}{\mu_2} u')^2) \sqrt{a_2 - u'} }{ (a_1 + (\frac{p}{\mu_2} u')^2) \sqrt{ u' - 1 } }
    \quad\mathrm{where}\quad u':= u_2\mu_2
    \label{eq:steady-states}
\end{equation}
which reveals that only $a_1$,$a_2$ and the ratio between half-saturation and degradation $\frac{k}{\mu_1}$ affect the solutions to this equation, and hence the locations of the bifurcations. Figure \ref{fig:two-state-optima} reveals the results for 800 optimisation runs with 200 ADAM iterations. 98\% of runs converged onto one of the two clusters: mutual activation and mutual inhibition regimes, examples of which are shown as insets.
\begin{figure}
\centering
%\includegraphics[width=11cm]{two-state-optima}
\includegraphics[width=11cm]{two-state-optima.png}
\caption{Optimal parameter estimates $\theta^*$ for the two state model \eqref{eq:two-state} for the targets $\targets=\{4,5\}$ reveal two clusters of qualitatively different regimes: mutual activation and mutual inhibition. }
\label{fig:two-state-optima}
\end{figure}
